A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-04-25 tidyindex.qmd
further tidying of intro

To be able to test different decision choices for an index, systematically and statistically, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component. We call this process the *index pipeline*, which are the steps of the data analysis pipeline for index construction. Such a decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing among indexes, calculating index uncertainty, and assessing index robustness. In addition it provides an unambiguous documentation for the index definition, facilitating reproducibility of results.

In this work, we provide statistical and computational methods for developing a data pipeline framework to construct and customize indexes using data. The proposed pipeline comprises various modules, including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. Given the decisions analysts need to make when combining multivariate data into indexes, the proposed pipeline enables the evaluation of how the specific choice can affect the index, as well as how the index may appear under alternative options. Furthermore, uncertainty calculation can also flow through the pipeline, providing the index with confidence measures.

>>>>

To be able to test different decision choices for an index, systematically and statistically, the index needs to be broken down into its fundamental building blocks to analyze the contribution and effect of each component. We call this process the *index pipeline*, which are the steps of the data analysis pipeline for index construction. Such a decomposition of index components provides the means to standardize index construction via a pipeline and offers benefits for comparing versions of indexes, calculating index uncertainty, and assessing index robustness. It also provides clear recipes for the index definition, facilitating reproducibility of results.

Here we detail the statistical and computational methods for developing a data pipeline framework to construct and customize indexes using data. The pipeline comprises various modules, including temporal and spatial aggregation, variable transformation and combination, distribution fitting, benchmark setting, and index communication. When combining multivariate data into indexes, the pipeline enables the evaluation of how any particular combination can affect the index. Uncertainty calculation can also flow through the pipeline to provide an index with confidence intervals. The pipeline also fits neatly into current tidy data workflows and data visualisation.

### 2024-04-25 tidyindex.tex
further tidying of intro

To be able to test different decision choices for an index,
systematically and statistically, the index needs to be broken down into
its fundamental building blocks to analyze the contribution and effect
of each component. We call this process the \emph{index pipeline}, which
are the steps of the data analysis pipeline for the index construction.
Such decomposition of index components provides the means to standardize
index construction via a pipeline and offers benefits for comparing
among indexes, calculating index uncertainty, and assessing index
robustness.

In this work, we provide statistical and computational methods for
developing a data pipeline framework to construct and customize indexes
using data. The proposed pipeline comprises various modules, including
temporal and spatial aggregation, variable transformation and
combination, distribution fitting, benchmark setting, and index
communication. Given the decisions analysts need to make when combining
multivariate data into indexes, the proposed pipeline enables the
evaluation of how the specific choice can affect the index, as well as
how the index may appear under alternative options. Furthermore,
uncertainty calculation can also flow through the pipeline, providing
the index with confidence measures.

@@

The state of Queensland in Australia frequently experiences natural
disaster events such as flood and drought, which can significantly
impact its agricultural industry. This example uses daily data from
Global Historical Climatology Network Daily (GHCND), aggregated into
monthly precipitation, to compute two drought indexes -- SPI and SPEI --
at various time scales and fitted distributions, for 29 stations in the
state of Queensland in Australia, spanning from January 1990 to April
2022. This example showcases the basic calculation of indexes with
different parameter specifications within the pipeline framework. The
dataset used in this example is available in the \texttt{tidyindex}
package as \texttt{queensland} and blow prints the first few rows of the
data:

@@

\begingroup\fontsize{11}{13}\selectfont

\begin{longtable}[t]{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}

\toprule
\textbf{Variable} & \textbf{V-wgt} & \textbf{Dimension} & \textbf{D-wgt} & \textbf{wgt}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25 & 0.050\\
Wage equality for similar work & 0.310 &  &  & 0.078\\
Estimated earned income & 0.221 &  &  & 0.055\\
Legislators senior officials and managers & 0.149 &  &  & 0.037\\
Professional and technical workers & 0.121 &  &  & 0.030\\
\addlinespace
Literacy rate & 0.191 & Education & 0.25 & 0.048\\
Enrolment in primary education & 0.459 &  &  & 0.115\\
Enrolment in secondary education & 0.230 &  &  & 0.058\\
Enrolment in tertiary education & 0.121 &  &  & 0.030\\
Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
\addlinespace
Healthy life expectancy & 0.307 &  &  & 0.077\\
Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
Women in ministerial positions & 0.247 &  &  & 0.062\\
Years with female head of state & 0.443 &  &  & 0.111\\
\bottomrule

@@

\tabularnewline

\end{longtable}

\endgroup{}

>>>>

To be able to test different decision choices for an index,
systematically and statistically, the index needs to be broken down into
its fundamental building blocks to analyze the contribution and effect
of each component. We call this process the \emph{index pipeline}, which
are the steps of the data analysis pipeline for index construction. Such
a decomposition of index components provides the means to standardize
index construction via a pipeline and offers benefits for comparing
versions of indexes, calculating index uncertainty, and assessing index
robustness. It also provides clear recipes for the index definition,
facilitating reproducibility of results.

Here we detail the statistical and computational methods for developing
a data pipeline framework to construct and customize indexes using data.
The pipeline comprises various modules, including temporal and spatial
aggregation, variable transformation and combination, distribution
fitting, benchmark setting, and index communication. When combining
multivariate data into indexes, the pipeline enables the evaluation of
how any particular combination can affect the index. Uncertainty
calculation can also flow through the pipeline to provide an index with
confidence intervals. The pipeline also fits neatly into current tidy
data workflows and data visualisation.

@@

The state of Queensland in Australia frequently experiences natural
disaster events such as flood and drought, which can significantly
impact its agricultural industry. This example uses daily data from
Global Historical Climatology Network Daily (GHCND), aggregated into
monthly precipitation, to compute two drought indexes -- SPI and SPEI --
at various time scales and fitted distributions, for 29 stations in the
state of Queensland in Australia, spanning from January 1990 to April
2022. This example showcases the basic calculation of indexes with
different parameter specifications within the pipeline framework. The
dataset used in this example is available in the \texttt{tidyindex}
package as \texttt{queensland} and below we show the first few rows of
the data:

@@

\begin{table}

\centering\begingroup\fontsize{11}{13}\selectfont

\begin{tabular}{>{\raggedright\arraybackslash}p{17.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedright\arraybackslash}p{3.5em}>{\raggedleft\arraybackslash}p{4em}>{\raggedleft\arraybackslash}p{3em}}
\toprule
\textbf{Variable} & \textbf{V-wgt} & \textbf{Dimension} & \textbf{D-wgt} & \textbf{wgt}\\
\midrule
Labour force participation & 0.199 & Economy & 0.25 & 0.050\\
Wage equality for similar work & 0.310 &  &  & 0.078\\
Estimated earned income & 0.221 &  &  & 0.055\\
Legislators senior officials and managers & 0.149 &  &  & 0.037\\
Professional and technical workers & 0.121 &  &  & 0.030\\
\hline
Literacy rate & 0.191 & Education & 0.25 & 0.048\\
Enrolment in primary education & 0.459 &  &  & 0.115\\
Enrolment in secondary education & 0.230 &  &  & 0.058\\
Enrolment in tertiary education & 0.121 &  &  & 0.030\\
\hline
Sex ratio at birth & 0.693 & Health & 0.25 & 0.173\\
Healthy life expectancy & 0.307 &  &  & 0.077\\
\hline
Women in parliament & 0.310 & Politics & 0.25 & 0.078\\
Women in ministerial positions & 0.247 &  &  & 0.062\\
Years with female head of state & 0.443 &  &  & 0.111\\
\bottomrule
\end{tabular}
\endgroup{}

@@

\end{table}%
